Improved Particle Approximations \\ to the Joint Smoothing Distribution \\ Using Markov Chain Monte Carlo

### 2012-06-05 Paper/complete/revisions_MCMCsmoother.tex
Post-review revisions and MATLAB

In a particle filter, the filtering distribution is approximated by a set of weighted samples (or ``particles'') drawn from that distribution.
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(x_{k}|y_{1:k}) & = & \sum_i w_k^{(i)} \delta_{x_k^{(i)}}(x_k)
\end{IEEEeqnarray}

@@

Similar ideas apply to the problem of smoothing. In the linear Gaussian case, the entire state sequence, $x_{1:K}$, may be estimated in closed-form using the a Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}. This works by modifying the ordinary Kalman filter results in a backward processing pass through the state sequence. For nonlinear or non-Gaussian models, it is possible again to resort to numerical methods. Now, particles are used to approximate the joint smoothing distribution.
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(x_{1:K}|y_{1:K}) & = & \sum_i w_K^{(i)} \delta_{x_{1:K}^{(i)}}(x_{1:K})
\end{IEEEeqnarray}

@@

The proposal for $x_k$, $q(x_{x}|x_{k-1})$, is often called the ``importance density''. Common choices for this include the model transition density (\ref{eq:state_proc}), and the ``optimal importance density'' \cite{Doucet2000a}. The proposal density for $x_{k-1}$, $q(x_{k-1})$, is constructed from the particles of the filtering approximation from the previous time step. The choice of weights and sampling procedure for $x_{k-1}$ determines what from of ``resampling'' is to be used. Here we use arbitrary weights for generality.
%
\begin{IEEEeqnarray}{rCl}
q(x_{k-1}) & = & \sum_i v_{k-1}^{(i)} \delta_{x_{k-1}^{(i)}}(x_{k-1})
\end{IEEEeqnarray}

@@

Next, the particles are assigned an importance weight according to the ratio of the target and the proposal densities.
%
\begin{IEEEeqnarray}{rCl}
w_{k}^{(i)} & =       & \frac{ p(x_{k-1}^{(i)}, x_k^{(i)}|y_{1:k}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) q(x_{k-1}^{(i)}) } \nonumber \\
            & \propto & \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) p(x_{k-1}^{(i)}|y_{1:k-1}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) q(x_{k-1}^{(i)}) } \nonumber \\
            & \approx & \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }
\end{IEEEeqnarray}

Normalisation is enforced by scaling the weights so that they sum to 1. Finally, $x_{k-1}$ is marginalised from the distribution by simply discarding the $x_{k-1}^{(i)}$ values from each particle.

The particle filter is summarised in Fig.~\ref{alg:PF}.

%\begin{algorithm}
\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
	\FOR{$k = 1$ \TO $K$}
 		\FOR{$i = 1$ \TO $N_F$}
 		  \STATE Resample last state $x_{k-1}^{(i)} \sim \sum_j v_{k-1}^{(j)} \delta_{x_{k-1}^{(j)}}(x_{k-1})$.
 			\STATE Sample new state $x_k^{(i)} \sim q(x_k|x_{k-1}^{(i)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 			\STATE Discard $x_{k-1}^{(i)}$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
\end{algorithmic}
}}
\caption{Particle filter algorithm}
\label{alg:PF}
\end{figure}
%\end{algorithm}

@@

%\begin{algorithm}
\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
 	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
 	\FOR{$k = 1$ \TO $K$}
 		\FOR{$i = 1$ \TO $N_F$}
 			\STATE Resample history $x_{1:k-1}^{(i)} \sim \sum_j v_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1})$.
 			\STATE Sample new state $x_k^{(i)} \sim q(x_k|x_{k-1}^{(i)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
\end{algorithmic}
}}
\caption{Filter-smoother algorithm}
\label{alg:FS}
\end{figure}
%\end{algorithm}

An improved particle smoothing algorithm was presented in \cite{Godsill2004}. This works by constructing new particles from the smoothing distribution by resampling from the filtering approximations. The sampling procedure exploits the following factorisation of the joint distribution.
%
\begin{IEEEeqnarray}{rCl}
p(x_{1:K}|y_{1:K}) & = & p(x_K|y_{1:K}) \prod_{k=1}^{K-1} p(x_k|x_{k+1}, y_{1:K}) \label{eq:smoothing_factorisation}
\end{IEEEeqnarray}

@@

Substituting the filtering approximation into this expression yields a particle distribution which may be sampled easily.
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(x_k|\tilde{x}_{k+1}, y_{1:K}) = \sum_i  \tilde{w}_k^{(i)} \delta_{x_{k}^{(i)}}(x_{k}) \label{eq:backward_conditional_filter} \\
\tilde{w}_k^{(i)} \propto w_k^{(i)} p(\tilde{x}_{k+1}|x_k^{(i)}) \label{eq:DBRS_weights}
\end{IEEEeqnarray}

@@

%\begin{algorithm}
\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
 	\STATE Run a particle filter to approximate $p(x_k|y_{1:k})$ for each $k$ and store the resulting particles $\{x_k^{(i)}, w_k^{(i)}\}$.
	\FOR{$i = 1$ \TO $N_S$}
		\STATE Sample $\tilde{x}_{K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{K}^{(j)}}(x_{K})$.
		\FOR{$k = K-1$ \TO $1$}
			\FOR{$j = 1$ \TO $N_F$}
				\STATE Calculate weight $\tilde{w}_k^{(j)} = w_k^{(j)} p(\tilde{x}_{k+1}|x_k^{(j)})$.
			\ENDFOR
			\STATE Sample $\tilde{x}_{k}^{(i)} \sim \sum_j \tilde{w}_k^{(j)} \delta_{x_{k}^{(j)}}(x_{k})$.
		\ENDFOR
	\ENDFOR
\end{algorithmic}
}}
\caption{Direct backward-resampling smoother algorithm}
\label{alg:DBRS}
\end{figure}
%\end{algorithm}

A method for drawing samples from the joint smoothing distribution, $p(x_{1:K}|y_{1:K})$, is included in \cite{Briers2010}. The principal topic of this paper is a two-filter smoother for estimation of the marginal smoothing distributions, $p(x_{k}|y_{1:K})$, by combination of the results from one forward and one backward particle filter. Particles from the joint distribution may then be generated by resampling from either the forward or backward filters (or some combination) in the manner of the backward-resampling smoother outlined above.%algorithm~\ref{alg:DBRS}.

The complexity of both the backward-resampling smoother of \cite{Godsill2004} and the two-filter variant of \cite{Briers2010} is $\mathcal{O}(N_F \times N_S \times K)$, where $N_F$ is the number of filter particles, $N_S$ the number of smoother particles, and $K$ the number of time steps. In \cite{Douc2009}, a theoretically linear-cost scheme using rejection sampling is outlined, but in practice this appears to be slower than the standard algorithms. Methods for approximation of such algorithms with lower complexity are discussed in \cite{Klaas2006}. However, computational cost is often still prohibitive, requiring either $N_S$ or $N_F$ to be held very low.

In the algorithms described so far, no new states are sampled during the smoothing stage; there is only a resampling of the filtering distributions. This limits the support of the smoother to the states that appear in the filtering particles. In \cite{Fearnhead2010}, an IS-based algorithm for estimation of the marginal smoothing distributions is described in which new states are sampled from a proposal, allowing full support. In addition, this algorithm achieves $\mathcal{O}( N_S \times K)$ complexity, linear in the number of particles.

%In section~\ref{sec:mcmc_smoother}, we introduce an MCMC approach for drawing samples from the joint smoothing distribution with an improved computational complexity over the direct backward-resampling smoother. In section~\ref{sec:new_state_smoother}, we combine this procedure with the new-proposal method used in \cite{Fearnhead2010} to generate joint smoothing distribution particles with full support.

@@

A valid proposal may be constructed using the particles of the filter-smoother approximation with any arbitrary weights.
%
\begin{IEEEeqnarray}{rCl}
q(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) & = & \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})
\end{IEEEeqnarray}

If the current state of the chain is $x_{1:k}^{(m)}$, and a new sample is drawn from the proposal $x_{1:k}^{*} \sim q(x_{1:k}|\tilde{x}_{k+1}, y_{1:K})$, then it should be accepted with the following probability:
%
\begin{IEEEeqnarray}{rCl}
\alpha &=& \min \Bigg[ 1, \frac{ \hat{p}(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) }{ \hat{p}(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) } \Bigg] \nonumber \\
       &=& \min \Bigg[1, \frac{ w_k^{*} \tilde{v}_k^{(m)} }{  w_k^{(m)} \tilde{v}_k^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) }  \Bigg]  . \label{eq:MCMC-BRS_ap}
\end{IEEEeqnarray}

Note that $w_k^{*}$ indicates the filtering particle weight of the proposed new state, $x_{1:k}^{*}$.

The final state of each chain is used as a sample from the required conditional distribution.

The complete MCMC backward-resampling smoother is summarised in Fig.~\ref{alg:MCMC-BRS}.

%\begin{algorithm}
\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$ and store the resulting particles $\{x_{1:k}^{(i)}, w_k^{(i)}\}$.
	\FOR{$i = 1$ \TO $N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1$ \TO $1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1$ \TO $M$}
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})$.
				\STATE With probability $\alpha$ specified by (\ref{eq:MCMC-BRS_ap}),\\ $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{1:k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
}}
\caption{MCMC backward-resampling smoother algorithm}
\label{alg:MCMC-BRS}
\end{figure}
%\end{algorithm}

@@

%The method described in \cite{Fearnhead2010} uses a forward and a backward particle filter. To develop a marginal smoothing approximation at time step $k$, particles are resampled from the forward filter at step $k-1$, and from the backward filter and step $k+1$. A new state for step $k$ is then sampled from a continuous proposal distribution conditional on the previous and next states, and an importance weight is assigned. It is undesirable to use importance sampling for generating particles from the joint sampling distribution, as the required resampling is liable to lead to a loss of particle diversity similar to that suffered by the filter-smoother.

@@

The proposed new method proceeds in a similar manner to the MCMC backward-resampling algorithm. However, this time only $x_{1:k-1}$ is resampled from the particle filter-smoother approximation from time step $k-1$. The current state, $x_k$, is then sampled afresh from a new proposal distribution. Thus the factorised proposal may be written as follows:
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ q(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) = q(x_{1:k-1}|y_{1:k-1}) q(x_{k}|x_{k-1}, \tilde{x}_{k+1}, y_{k}) } \nonumber \\
                                    & = & \sum_j \tilde{v}_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1}) q(x_{k}|x_{k-1}^{(j)}, \tilde{x}_{k+1}, y_{k})  .
\end{IEEEeqnarray}

The target distribution is also factorised.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) = p(x_{1:k}|\tilde{x}_{k+1}, y_{1:k}) } \nonumber \\
                                    & \propto & p(\tilde{x}_{k+1}|x_k) p(x_k|x_{k-1}) p(y_k|x_x) p(x_{1:k-1}|y_{1:k-1})
\end{IEEEeqnarray}

Using the filter-smoother approximation for $p(x_{1:k-1}|y_{1:k-1})$, the MH acceptance probabilities may then be calculated.
%
\begin{IEEEeqnarray}{rCl}
\alpha &=& \min \Bigg[ 1, \frac{ \hat{p}(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) }{ \hat{p}(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) } \Bigg] \nonumber \\
 &=& \min \Bigg[ 1, \frac{ w_{k-1}^{*} p(\tilde{x}_{k+1}|x_k^{*}) p(x_k^{*}|x_{k-1}^{*}) p(y_k|x_k^{*}) }{ w_{k-1}^{(m)} p(\tilde{x}_{k+1}|x_k^{(m)}) p(x_k^{(m)}|x_{k-1}^{(m)}) p(y_k|x_k^{(m)}) } \nonumber \\
 & & \times \frac{ \tilde{v}_{k-1}^{(m)} q(x_k^{(m)}|x_{k-1}^{(m)},\tilde{x}_{k+1},y_k) }{ \tilde{v}_{k-1}^{*} q(x_k^{*}|x_{k-1}^{*},\tilde{x}_{k+1},y_k) } \Bigg] \label{eq:MCMC-BSS_ap}
% \frac{ w_{k-1}^{*} \tilde{v}_{k-1}^{(m)} }{  w_{k-1}^{(m)} \tilde{v}_{k-1}^{*} } \times \nonumber \\
% \IEEEeqnarraymulticol{3}{l}{\frac{ p(\tilde{x}_{k+1}|x_k^{*}) p(x_k^{*}|x_{k-1}^{*}) p(y_k|x_k^{*}) q(x_k^{(m)}|x_{k-1}^{(m)},\tilde{x}_{k+1},y_k) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) p(x_k^{(m)}|x_{k-1}^{(m)}) p(y_k|x_k^{(m)}) q(x_k^{*}|x_{k-1}^{*},\tilde{x}_{k+1},y_k) } } .
\end{IEEEeqnarray}

%As noted in \cite{Fearnhead2010}, the optimal choice of proposal is the conditional posterior state distribution,

%\begin{IEEEeqnarray}{rCl}
%q(x_k|x_{k-1},x_{k+1},y_k) & = & p(x_k|x_{k-1},x_{k+1},y_k)  .
%\end{IEEEeqnarray}

The complete MCMC backward-sampling smoother is summarised in Fig.~\ref{alg:MCMC-BSS}.

>>>>

In a particle filter, the filtering distribution is approximated by a set of weighted samples (or ``particles'') drawn from that distribution,
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(x_{k}|y_{1:k}) & = & \sum_i w_k^{(i)} \delta_{x_k^{(i)}}(x_k)     .
\end{IEEEeqnarray}

@@

Similar ideas apply to the problem of smoothing. In the linear Gaussian case, the entire state sequence, $x_{1:K}$, may be estimated in closed-form using the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}. This works by modifying the ordinary Kalman filter results in a backward processing pass through the state sequence. For nonlinear or non-Gaussian models, it is possible again to resort to numerical methods. Now, particles are used to approximate the joint smoothing distribution,
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(x_{1:K}|y_{1:K}) & = & \sum_i w_K^{(i)} \delta_{x_{1:K}^{(i)}}(x_{1:K})     .
\end{IEEEeqnarray}

@@

The proposal for $x_k$, $q(x_{x}|x_{k-1})$, is often called the ``importance density''. Common choices for this include the model transition density (\ref{eq:state_proc}), and the ``optimal importance density'' \cite{Doucet2000a}. The proposal density for $x_{k-1}$, $q(x_{k-1})$, is constructed from the particles of the filtering approximation from the previous time step. The choice of weights and sampling procedure for $x_{k-1}$ determines what from of ``resampling'' is to be used. Here we use arbitrary weights for generality,
%
\begin{IEEEeqnarray}{rCl}
q(x_{k-1}) & = & \sum_i v_{k-1}^{(i)} \delta_{x_{k-1}^{(i)}}(x_{k-1})     .
\end{IEEEeqnarray}

@@

Next, the particles are assigned an importance weight according to the ratio of the target and the proposal densities,
%
\begin{IEEEeqnarray}{rCl}
w_{k}^{(i)} & =       & \frac{ p(x_{k-1}^{(i)}, x_k^{(i)}|y_{1:k}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) q(x_{k-1}^{(i)}) } \nonumber \\
            & \propto & \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) p(x_{k-1}^{(i)}|y_{1:k-1}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) q(x_{k-1}^{(i)}) } \nonumber \\
            & =       & \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }     .
\end{IEEEeqnarray}

Normalisation is enforced by scaling the weights so that they sum to 1. Finally, $x_{k-1}$ is marginalised from the distribution by simply discarding the $x_{k-1}^{(i)}$ values from each particle. The particle filter is summarised in Fig.~\ref{alg:PF}.

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
	\FOR{$k = 1$ \TO $K$}
 		\FOR{$i = 1$ \TO $N_F$}
 		  \STATE Resample last state $x_{k-1}^{(i)} \sim \sum_j v_{k-1}^{(j)} \delta_{x_{k-1}^{(j)}}(x_{k-1})$.
 			\STATE Sample new state $x_k^{(i)} \sim q(x_k|x_{k-1}^{(i)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 			\STATE Discard $x_{k-1}^{(i)}$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
\end{algorithmic}
}}
\caption{Particle filter algorithm}
\label{alg:PF}
\end{figure}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
 	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
 	\FOR{$k = 1$ \TO $K$}
 		\FOR{$i = 1$ \TO $N_F$}
 			\STATE Resample history $x_{1:k-1}^{(i)} \sim \sum_j v_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1})$.
 			\STATE Sample new state $x_k^{(i)} \sim q(x_k|x_{k-1}^{(i)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
\end{algorithmic}
}}
\caption{Filter-smoother algorithm}
\label{alg:FS}
\end{figure}

An improved particle smoothing algorithm was presented in \cite{Godsill2004}. This works by constructing new particles from the smoothing distribution by resampling from the filtering approximations. The sampling procedure exploits the following factorisation of the joint distribution,
%
\begin{IEEEeqnarray}{rCl}
p(x_{1:K}|y_{1:K}) & = & p(x_K|y_{1:K}) \prod_{k=1}^{K-1} p(x_k|x_{k+1}, y_{1:K}) \label{eq:smoothing_factorisation}     .
\end{IEEEeqnarray}

@@

Substituting the filtering approximation into this expression yields a particle distribution which may be sampled easily,
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(x_k|\tilde{x}_{k+1}, y_{1:K}) = \sum_i  \tilde{w}_k^{(i)} \delta_{x_{k}^{(i)}}(x_{k}) \label{eq:backward_conditional_filter} \\
\tilde{w}_k^{(i)} \propto w_k^{(i)} p(\tilde{x}_{k+1}|x_k^{(i)}) \label{eq:DBRS_weights}     .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
 	\STATE Run a particle filter to approximate $p(x_k|y_{1:k})$ for each $k$ and store the resulting particles $\{x_k^{(i)}, w_k^{(i)}\}$.
	\FOR{$i = 1$ \TO $N_S$}
		\STATE Sample $\tilde{x}_{K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{K}^{(j)}}(x_{K})$.
		\FOR{$k = K-1$ \TO $1$}
			\FOR{$j = 1$ \TO $N_F$}
				\STATE Calculate weight $\tilde{w}_k^{(j)} = w_k^{(j)} p(\tilde{x}_{k+1}|x_k^{(j)})$.
			\ENDFOR
			\STATE Sample $\tilde{x}_{k}^{(i)} \sim \sum_j \tilde{w}_k^{(j)} \delta_{x_{k}^{(j)}}(x_{k})$.
		\ENDFOR
	\ENDFOR
\end{algorithmic}
}}
\caption{Direct backward-resampling smoother algorithm}
\label{alg:DBRS}
\end{figure}

A method for drawing samples from the joint smoothing distribution, $p(x_{1:K}|y_{1:K})$, is included in \cite{Briers2010}. The principal topic of this paper is a two-filter smoother for estimation of the marginal smoothing distributions, $p(x_{k}|y_{1:K})$, by combination of the results from one forward and one backward particle filter. Particles from the joint distribution may then be generated by resampling from either the forward or backward filters (or some combination) in the manner of the backward-resampling smoother outlined above.

The complexity of both the backward-resampling smoother of \cite{Godsill2004} and the two-filter variant of \cite{Briers2010} is $\mathcal{O}(N_F \times N_S \times K)$, where $N_F$ is the number of filter particles, $N_S$ the number of smoother particles, and $K$ the number of time steps. In \cite{Douc2009}, a theoretically linear-cost scheme using rejection sampling is outlined, but in practice this often appears to be slower than the standard algorithms due to high rejection rates. Methods for reducing the computational complexity of smoothing algorithms are also discussed in \cite{Klaas2006}. However, these focus on approximating a sum of kernel values over a set of particles, such as those which arise in forward-backward smoothing \cite{Doucet2000a} and two-filter smoothing \cite{Briers2010} for estimation of the marginal smoothing distribution. They are not applicable for the joint smoothing procedures.

In the algorithms described so far, no new states are sampled during the smoothing stage; there is only a resampling of the filtering distributions. This limits the support of the smoother to the states that appear in the filtering particles. In \cite{Fearnhead2010}, an IS-based algorithm for estimation of the marginal smoothing distributions is described in which new states are sampled from a proposal density, thus recovering the full support. In addition, this algorithm achieves $\mathcal{O}( N_S \times K)$ complexity, i.e. linear in the number of particles.

@@

A valid proposal may be constructed using the particles of the filter-smoother approximation with any arbitrary weights,
%
\begin{IEEEeqnarray}{rCl}
q(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) & = & \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})     .
\end{IEEEeqnarray}

If the current state of the chain is $x_{1:k}^{(m)}$, and a new sample is drawn from the proposal $x_{1:k}^{*} \sim q(x_{1:k}|\tilde{x}_{k+1}, y_{1:K})$, then it should be accepted with the following probability,
%
\begin{IEEEeqnarray}{rCl}
\alpha &=& \min \Bigg[ 1, \frac{ \hat{p}(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) }{ \hat{p}(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) } \Bigg] \nonumber \\
       &=& \min \Bigg[1, \frac{ w_k^{*} \tilde{v}_k^{(m)} }{  w_k^{(m)} \tilde{v}_k^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) }  \Bigg]  . \label{eq:MCMC-BRS_ap}
\end{IEEEeqnarray}

Note that $w_k^{*}$ indicates the filtering particle weight of the proposed new state, $x_{1:k}^{*}$. The final state of each chain is used as a sample from the required conditional distribution. The complete MCMC backward-resampling smoother is summarised in Fig.~\ref{alg:MCMC-BRS}.

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$ and store the resulting particles $\{x_{1:k}^{(i)}, w_k^{(i)}\}$.
	\FOR{$i = 1$ \TO $N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1$ \TO $1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1$ \TO $M$}
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})$.
				\STATE With probability $\alpha$ specified by (\ref{eq:MCMC-BRS_ap}),\\ $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{1:k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
}}
\caption{MCMC backward-resampling smoother algorithm}
\label{alg:MCMC-BRS}
\end{figure}

@@

@@

The proposed new method proceeds in a similar manner to the MCMC backward-resampling algorithm. However, this time only $x_{1:k-1}$ is resampled from the particle filter-smoother approximation from time step $k-1$. The current state, $x_k$, is then sampled afresh from a new proposal distribution. Thus the factorised proposal may be written as follows,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ q(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) = q(x_{1:k-1}|y_{1:k-1}) q(x_{k}|x_{k-1}, \tilde{x}_{k+1}, y_{k}) } \nonumber \\
                                    & = & \sum_j \tilde{v}_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1}) q(x_{k}|x_{k-1}^{(j)}, \tilde{x}_{k+1}, y_{k})  .
\end{IEEEeqnarray}

The target distribution is also factorised,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) = p(x_{1:k}|\tilde{x}_{k+1}, y_{1:k}) } \nonumber \\
                                    & \propto & p(\tilde{x}_{k+1}|x_k) p(x_k|x_{k-1}) p(y_k|x_x) p(x_{1:k-1}|y_{1:k-1})     .
\end{IEEEeqnarray}

Using the filter-smoother approximation for $p(x_{1:k-1}|y_{1:k-1})$, the MH acceptance probabilities may then be calculated,
%
\begin{IEEEeqnarray}{rCl}
\alpha &=& \min \Bigg[ 1, \frac{ \hat{p}(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) }{ \hat{p}(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) } \Bigg] \nonumber \\
 &=& \min \Bigg[ 1, \frac{ w_{k-1}^{*} p(\tilde{x}_{k+1}|x_k^{*}) p(x_k^{*}|x_{k-1}^{*}) p(y_k|x_k^{*}) }{ w_{k-1}^{(m)} p(\tilde{x}_{k+1}|x_k^{(m)}) p(x_k^{(m)}|x_{k-1}^{(m)}) p(y_k|x_k^{(m)}) } \nonumber \\
 & & \times \frac{ \tilde{v}_{k-1}^{(m)} q(x_k^{(m)}|x_{k-1}^{(m)},\tilde{x}_{k+1},y_k) }{ \tilde{v}_{k-1}^{*} q(x_k^{*}|x_{k-1}^{*},\tilde{x}_{k+1},y_k) } \Bigg] \label{eq:MCMC-BSS_ap}     .
\end{IEEEeqnarray}

The complete MCMC backward-sampling smoother is summarised in Fig.~\ref{alg:MCMC-BSS}. The computational complexity of this algorithm is $\mathcal{O}(M \times N_S \times K)$, the same as the MCMC resampling smoother, although it is likely take more time to run because of the additional sampling operation and probability calculations. In contrast with the MCMC resampling method, this smoother targets the exact posterior distribution for the current state at time $t_n$, rather than a particle approximation. Consequently, the performance is expected to be improved relative to the resampling-based methods in terms of both error and particle diversity.

We note that this new scheme bears a resemblance to the improvement strategy for approximations to the joint smoothing distribution published recently in \cite{Dubarry2011}. Their algorithm iteratively applies MH steps to a low-diversity particle approximation, such as that from the filter-smoother. Our algorithm differs in the fact that it allows simultaneous changes to the particle state history, $x_{1:n-1}$, as well as the current state. Furthermore, for each smoothing particle, we run a Markov chain at each time step, using one sweep backwards through time, rather than multiple sweeps with a single MH step at each time point.
